---

title: "Spatial trends in crime rates across pomeranian counties in 2018"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    bibliography: bibliography.bib 
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```


# *Spatial Trends in Property Crimes: A Focus on crimes against property across Poland's  Pomeranian voivodeship*

## INTRODUCTION

The study aims to analyze spatial trends in property crimes, in Poland in regards to the Pomeranian voivodeship and it’s counties. Further examination explores how socio-economic factors---like unemployment rate, social services, average gross monthly salary, population density --- affect crime rates, considering geographic contexts. The study takes data from 2018.

## I. DATA DESCRIPTION

The database comprises data on the number of ascertained crimes in 2018 across Poland's Pomeranian voivodeship and it's counties.

**Crime ascertained** is a crime concluded with confirmation of the occurrence of the prohibited act.

The database contain only the crimes committed by an adult.

[Note:]{.underline} Polish law indicates that theft with a value not exceeding 500 PLN is considered a misdemeanor, not a crime. Reporting such theft will not be considered or processed, so there is no point in reporting them. Such a situation causes a disturbance between the actual amount of thefts and the recorded ones, since people has no incentive to report a theft under 500 PLN. This aspect was taken into consideration during the examination.

We created a variable "crimesrate" by examining the shares of particular county in the general sum of crimes in the pomeranian voivodeship making it as a rate. 

## II. DATA VISUALIZATION

```{r}
install.packages(sf)
install.packages(rgugik)
install.packages(bdl)
install.packages(elevatr)
install.packages(terra)
install.packages("ggplot2")
install.packages("tidyverse")
install.packages("giscoR")
install.packages("mapsf")
install.packages("spatialreg")
install.packages("hglm")
install.packages("mgcv")
install.packages("tmap")
install.packages("lmtest")
install.packages("zoo")
install.packages("plot.matrix")
install.packages("viridis")
install.packages("sphet")
install.packages("spdep")
install.packages("spatialreg")
install.packages("grid")
install.packages("tinyplot", repos = "https://grantmcdermott.r-universe.dev")
install.packages("RSpectra")
install.packages("igraph")

```

```{r}
library(rgugik)
library(sf)
library(bdl)
library(giscoR)
library(mapsf)
library(sp)
library(spdep)
library(spatialreg)
library(hglm)
library(mgcv)
library(tmap)
library(lmtest)
library(zoo)
library(plot.matrix)
library(viridis)
library(sphet)
library(spdep)
library(spatialreg)
library(tinyplot)
library(grid)
library(RSpectra)
library(igraph)
library(knitcitations)
```
### CREATING THE DATABASE 

```{r}
substring(county_names$TERYT, 1, 2) -> vopom
county_names |> subset(subset = vopom == "22") -> pom01
```

```{r}
borders_get(TERYT = pom01$TERYT) -> wpom
ID <- as.integer("0012345") 
str(ID)
formatC(ID, format="d", width=7, flag="0")
dim(wpom)
wpom |> sapply(class) |> str()
str(wpom$TERYT)
str(pom01)
```
First, to collect the spatial data and extract boarders across pomeranian voivodeship, we are using rgugik package that include open data from resources of Polish Head Office of Geodesy and Cartography.


```{r}
any(is.na(match(wpom$TERYT, pom01$TERYT)))
```

```{r}
wpom |> merge(pom01, by = "TERYT") -> wpom1 
wpom1
```
Merging the pom object containing the "TERYT" keys and municipality names, with pom_sf, including the "TERYT" keys and municipality boundaries. 


```{r}
all <- readxl::read_excel("all_variables.xlsx")
sapply(all, class)
names(all)
wpom1 |> merge(all, by.x = "TERYT",
 by.y = "Code") -> wpom2
wpom2
```
Now we add socio-economic data about unemployment, social services, average gross monthly salary, population density in pomeranian voivodeship, collected from GUS.
We merged the 'wpom1' spatial dataset with the dataset containing socio-economic factors, based on matching values in the "TERYT" column of 'wpom1' and the "Code" column of 'all'.


```{r}
wpom2 |> st_area() |>
 units::set_units("km2") -> wpom2$area_km2
wpom2 |> st_drop_geometry() |>
 subset(select = c(poplation, area_km2)) |>
 apply(1, function(x) x[1]/x[2]) -> wpom2$density
 wpom2
```
New dataset called 'wpom2' contains the merged data, showing information about municipalities within the Pomeranian Voivodeship enriched with additional variables.


### UNEMPLOYMENT RATE

We assume that there is direct proportional relationship between unemployment rate and crimes against property (the higher rate of unemployment, the higher amount of crimes against property).```

```{r}
library(mapview)
mapview(wpom2, zcol="unemp")
```

Interactive map of Pomeramia municipalities from rgugik, unemployment rate 
after all download administrative municipality boundaries using rgugik

```{r}
pom_gisco <- gisco_get_nuts(year="2021", resolution="01",
 spatialtype="RG", nuts_id="PL63")
pom_gisco_tm <- st_transform(pom_gisco, "EPSG:2180")
wpom3 <- st_intersection(wpom2, pom_gisco_tm)
```

```{r}
wpom2 |> st_geometry_type() |> table() -> tab5; tab5[tab5>0]
```


```{r}
wpom3 |> st_geometry_type() |> table() -> tab6; tab6[tab6>0]
```


```{r}
wpom3 |> st_cast("MULTIPOLYGON") -> pom3
```

```{r}
mapview(wpom3, zcol="unemp")
```

```{r}
plot(density(wpom3$unemp))
```

```{r}
mapsf::mf_map(wpom3, var="unemp", type="choro", breaks="geom",
 nbreaks=7)
```

### SOCIAL SERVICES USERS

A direct proportional relationship between social services users and crimes against property is assumed (the higher rate of social services users, the higher amount of crimes against property).

```{r}
mapview(wpom3, zcol="ssusers")
```

```{r}
plot(density(wpom3$ssusers))
```

```{r}
mapsf::mf_map(wpom3, var="ssusers", type="choro", breaks="geom",
 nbreaks=7)
```

### AVERAGE GROSS MONTHLY SALARY 

An inverse relationship between average gross monthly salary and crimes against property is assumed (the higher amount of average gross monthly salary, the lower amount of crimes against property).

```{r}
mapview(wpom3, zcol="avgsal")
```

```{r}
plot(density(wpom3$avgsal))
```

```{r}
mapsf::mf_map(wpom3, var="avgsal", type="choro", breaks="geom",
              nbreaks=7)
```

### POPULATION DENSITY 

It is assumed that the number of crimes committed is directly proportional to the population density of cities, meaning we expect a greater number of thefts in larger urban areas.

```{r}
mapview(wpom3, zcol="density")
```

```{r}
plot(density(wpom3$density))
```

```{r}
mapsf::mf_map(wpom3, var="density", type="choro", breaks="geom",
              nbreaks=7)
```

### CRIMES

```{r}
mapview(wpom3, zcol="crimesrate")
```

```{r}
plot(density(wpom3$ratecrimes))
```

```{r}
mapsf::mf_map(wpom3, var="crimesrate", type="choro", breaks="geom",
 nbreaks=7)
```
```{r}
cor(model.matrix(~ -1 + crimesrate + density + avgsal + unemp + ssusers,wpom3))
```
The strongest correlation is between 'crimesrate' and 'avgsal' - 0.6957, suggesting a moderately positive relationship. 
Also 'unemp' and 'ssusers' have a correlation coefficient of 0.6813, indicating a moderately positive relationship.
'density' and 'avgsal' have a strong positive correlation coefficient of 0.8042, indicating that areas with higher population density tend to have higher average salaries.
Crime rates is assumed to be negatively correlated with unemployment and ssusers, which is not very logical. 

```{r}
(wpom3 |> poly2nb(queen=FALSE,
 row.names=wpom3$TERYT) -> pom_rook_nb)
```

```{r}
(wpom3 |> poly2nb(queen=TRUE,
 row.names=wpom3$TERYT) -> pom_queen_nb)
```

```{r}
(wpom3 |> poly2nb(queen=FALSE, row.names=wpom3$TERYT,
 snap=0.00001) -> pom_rook_nb)
(wpom3 |> poly2nb(queen=TRUE, row.names=wpom3$TERYT,
 snap=0.00001) -> pom_queen_nb)
```
Number of regions: There are 20 regions/entities in the dataset.
Number of nonzero links: There are 78 pairwise connections between these regions. Nonzero links indicate spatial relationships where regions share a boundary or are otherwise considered neighbors.
Percentage nonzero weights: Nonzero weights account for 19.5% of all possible spatial relationships in the dataset. This indicates the density of spatial connections among the regions.
Average number of links: On average, each region has approximately 3.9 neighbors. This metric provides insight into the spatial connectivity and clustering patterns within the dataset.



Now we want to establish whether entities without neighbors truly have no neighbors and detect disjoint subgraphs within a spatial object. This is crucial for ensuring the accuracy of spatial analysis and modeling, as certain methods may produce unreliable results if observations are "dropped out" due to lacking neighbors or if the spatial structure is fragmented.

```{r}
geom <- st_geometry(wpom2) 
plot(geom, border="grey")
plot(pom_queen_nb, geom, add=TRUE)
```
```{r}
(geom |> st_point_on_surface() |> knearneigh(k=5) |>
 knn2nb(row.names=wpom2$TERYT) -> pom_1_nb)
```

```{r}
plot(geom, border="grey")
plot(pom_1_nb, geom, add=TRUE)
```

```{r}
pom_queen_nb |> nb2listw(style="W") -> pom_queen_lw
summary(pom_queen_lw)
```

```{r}
pom_queen_lw$weights |> sapply(sum) |> unique()
```

## Regression Model

```{r}
model <- lm(crimesrate ~ density + avgsal + unemp + ssusers, data = wpom3)
summary(model)
coefficients(model)
residuals(model)
```
The R-squared value of 0.5225 indicates that approximately 52.25% of the variability in crime rates is explained by the independent variables in the model.

Density: The coefficient for 'density' is close to zero and statistically insignificant (p-value = 0.998). This indicates that population density doesn't significantly influence crime rates in this model.
Avgsal (Average Salary): The coefficient for 'avgsal' is positive (0.0052), but statistically insignificant (p-value = 0.181), suggesting that average salary may not be a significant predictor of crime rates.
Unemp (Unemployment Rate): The coefficient for 'unemp' is negative (-0.0639), but statistically insignificant (p-value = 0.903), indicating that unemployment rate may not have a significant impact on crime rates.
Ssusers (Social Security Users): The coefficient for 'ssusers' is negative (-1.191), but statistically insignificant (p-value = 0.368), suggesting that the number of social security users may not be a significant determinant of crime rates.



## TEST OF SPACIAL AUTOCORRELATION

```{r}

lw <- nb2listw(pom_queen_nb, style = "W")

moran_test <- moran.test(wpom3$crimesrate, listw = lw, randomisation = TRUE, alternative = "two.sided")
print(moran_test)

```
The Moran's I test was conducted to examine whether there is spatial autocorrelation in property crimes across Poland's Pomeranian voivodeship. 
(Spatial autocorrelation refers to the tendency of similar values to cluster together in space.)

Since the p-value (0.2962) is greater than the conventional significance level of 0.05, we fail to reject the null hypothesis. This suggests that there is insufficient evidence to conclude that there is significant spatial autocorrelation in the variable crimes across the region. Therefore, there is no evidence of spatial clustering or spatial dependence in the crime rates based on the Moran's I test


###TEST OF RESIDUAL SPATIAL AUTOCORRELATION

```{r}

lm_null <- lm(crimesrate ~ 1, data = wpom3)

moran_test <- moran.test(residuals(lm_null), listw = lw, randomisation = FALSE, alternative = "two.sided")

print(moran_test)

```
The Moran's I test conducted on the residuals of the linear regression model indicates no significant spatial autocorrelation (p = 0.4692). Thus, there is no evidence to suggest that neighboring regions exhibit similar crime rates beyond what would be expected by chance, under the assumtion of normality 

```{r}
lm.morantest(lm_null, listw=lw, alternative="two.sided")
```
same result, no spatial autocorrelation in the residuals

```{r}
form_pom <- crimesrate ~ density + avgsal + unemp + ssusers
lm_obj <- lm(form_pom, data= wpom3)
moran.test(residuals(lm_obj), listw=lw,
 randomisation=FALSE, alternative="two.sided")
```
```{r}
lm.morantest(lm_obj, listw=lw, alternative="two.sided")
```

These results suggest that there is evidence of spatial autocorrelation in the residuals of the regression model under the assumption of normality at the specified level of significance. The p-value of 0.0351 is less than the common significance level of 0.05, indicating that we reject the null hypothesis of spatial randomness in favor of the alternative hypothesis of spatial autocorrelation. The negative Moran's I statistic suggests a negative spatial autocorrelation, indicating that neighboring regions tend to have dissimilar residuals. Therefore, the spatial structure should be considered in the regression analysis to account for the spatial autocorrelation in the residuals.





```{r}
lm.morantest.sad(lm_obj, listw=lw, alternative="two.sided")
```
The p-value of 0.04643 suggests that there is evidence to reject the null hypothesis of spatial randomness in favor of the alternative hypothesis of spatial autocorrelation at the significance level of 0.05. This indicates that there may be significant spatial clustering or dispersion in the residuals of the regression model.
The observed Moran's I statistic provides a measure of the spatial autocorrelation in the residuals, indicating whether neighboring regions tend to have similar or dissimilar residuals. In this case, the negative Moran's I statistic suggests a negative spatial autocorrelation, implying that neighboring regions tend to have dissimilar residuals.


## RAO'S SCORE

```{r}
summary(lm.RStests(lm_obj, listw=lw, test="all"))
```
```{r}
summary(SD.RStests(lm_obj, listw=lw, test="SDM"))
```
```{r}
summary(SD.RStests(lm_obj, listw=lw, test="SDEM"))
```
```{r}
SLX_obj <- lmSLX(form_pom, data=wpom3, listw=lw)
summary(lm.RStests(SLX_obj, listw=lw, test="all"))
```

Rao's score (also known as Lagrange multiplier) diagnostics for spatial dependence provide insights into the presence of spatial autocorrelation in the residuals of a regression model. Here are the key findings from the diagnostics:

RSerr (Residual Spatial Error): The statistic is 4.8395 with 1 degree of freedom and a p-value of 0.027815, indicating that there is evidence of spatial autocorrelation in the residuals. The significance level is denoted by a single asterisk (*), suggesting significance at the 0.05 level.
RSlag (Lag Spatial): The statistic is 7.5599 with 1 degree of freedom and a p-value of 0.005968, indicating strong evidence of spatial autocorrelation in the residuals. The significance level is denoted by two asterisks (**), suggesting significance at the 0.01 level.
adjRSerr (Adjusted Residual Spatial Error): The statistic is 0.059416 with 1 degree of freedom and a p-value of 0.807422, indicating no significant evidence of spatial autocorrelation in the residuals.
adjRSlag (Adjusted Lag Spatial): The statistic is 2.7798 with 1 degree of freedom and a p-value of 0.095461, suggesting borderline significance in detecting spatial autocorrelation in the residuals.
SARMA (Spatial Autoregressive Moving Average): The statistic is 7.6193 with 2 degrees of freedom and a p-value of 0.022156, indicating evidence of spatial autocorrelation in the residuals. The significance level is denoted by a single asterisk (*), suggesting significance at the 0.05 level.
In summary, the diagnostics indicate the presence of spatial autocorrelation in the residuals of the regression model, particularly highlighted by the significant findings in RSerr, RSlag, and SARMA tests. This suggests that spatial regression techniques may be more appropriate to account for spatial dependence in the data.



##SIMULTANEOUS AUTOREGRESSIVE MODELS

```{r}
sar1B <- spautolm(form_pom, data=wpom3, listw=lw, family="SAR")
summary(sar1B, Nagelkerke=TRUE)
```
```{r}
1/range(eigenw(lw))
```
```{r}
lw <- nb2listw(pom_queen_nb, style="W")
e <- eigenw(lw)
1/range(e)
```
```{r}
sar1 <- spautolm(form_pom, data=wpom3, listw=lw, family="SAR", control=list(pre_eig=e))
summary(sar1, Nagelkerke=TRUE)
```
```{r}
SEM <- errorsarlm(form_pom, data=wpom3, listw=lw, control=list(pre_eig=e))
summary(SEM, Nagelkerke=TRUE)
```

```{r}
all.equal(coef(SEM)[c(2:6, 1)], coef(sar1), tol=2e-07)
```

```{r}
Hausman.test(SEM)
```
## Estimation of spatial autoregressive models

```{r}
lw <- nb2listw(pom_queen_nb, style="W")
e <- eigenw(lw)
```


### Maximum likelihood estimation

```{r}
SEM <- errorsarlm(form_pom, data=wpom3, listw=lw, control=list(pre_eig=e), quiet=FALSE)
```
 
 
```{r}
SAC <- sacsarlm(form_pom, data=wpom3, listw=lw, control=list(pre_eig1=e, pre_eig2=e), llprof=40)
SAC_track <- capture.output(sac <- sacsarlm(form_pom, data=wpom3, listw=lw, control=list(pre_eig1=e, pre_eig2=e), quiet=FALSE))
c(SAC$rho, SAC$lambda)
c(SAC$rho/SAC$rho.se, SAC$lambda/SAC$lambda.se)
```
 
```{r}
m <- -matrix(SAC$llprof$ll, 40, 40)
con <- textConnection(SAC_track)
sac_track <- read.table(con, skip=14)
close(con)
contour(SAC$llprof$xseq, SAC$llprof$yseq, m, levels=quantile(c(m), seq(0,1,0.1)), xlab="rho", ylab="lambda", col="blue4")
abline(h=SAC$rho, v=SAC$lambda, lwd=3, col="grey")
lines(sac_track$V2, sac_track$V4, col="brown3")
```

### Bayesian estimation

```{r}
set.seed(12345)
SAC_bayes <- spBreg_sac(form_pom, data=wpom3, listw=lw, control=list(ndraw=20000L, nomit=2000L))
```

```{r}
summary(SAC_bayes[, c("rho", "lambda", "sige")])
```

```{r}
opar <- par(mfrow=c(2, 1)) 
plot(SAC_bayes[, c("rho", "lambda", "sige")], smooth=TRUE)
```

### Generalised method of moments estimation

```{r}
SAC <- spreg(form_pom, data=wpom3, listw=lw, model="sarar")
c(coef(SAC)[c("lambda", "rho"),], s2=as.numeric(SAC$s2))
```

## Trace-style impacts

```{r}
lw <- nb2listw(pom_queen_nb, style="W")
e <- eigenw(lw)
W <- as(lw, "CsparseMatrix")
trMat <- trW(W, type="mult")
```

```{r}
SLM <- lagsarlm(form_pom, data=wpom3, listw=lw, control=list(pre_eig=e))
```

## Eigenvectors and eigenvalues of graphs of relationships between cross-sectional observations

```{r}
lwB <- nb2listw(pom_queen_nb, style="B")
lwW <- nb2listw(pom_queen_nb, style="W")
```

```{r}
B0 <- as(lwB, "symmetricMatrix")
```
```{r}
g1 <- graph_from_adjacency_matrix(B0, mode="undirected")
c1 <- components(g1)
str(c1)
```
```{r}
str(n.comp.nb(pom_queen_nb))
```
```{r}
is_connected(g1)
```
```{r}
diameter(g1)
```
```{r}
str(g1d <- distances(g1))
```
```{r}
summary(c(g1d))
```
```{r}
head(sort(degree(g1), decreasing=TRUE))
```
```{r}
cunb <- card(pom_queen_nb)
names(cunb) <- attr(pom_queen_nb, "region.id")
head(sort(cunb, decreasing=TRUE))
```
```{r}
head(sort(page_rank(g1)$vector, decreasing=TRUE))
```
```{r}
head(sort(eigen_centrality(g1)$vector, decreasing=TRUE))
```
### Eigenvalue approach

```{r}
eB <- eigenw(lwB)
eW <- eigenw(similar.listw(lwW))
cat("Class:", class(eB), "min:", 1/min(eB), "max:", 1/max(eB), "\n")
??eigen
```

```{r}
eigW <- eigenw(similar.listw(lwW))
cat("Class:", class(eW), "min:", 1/min(eW), "max:", 1/max(eW), "\n")
```

```{r}
oopar <- par(mfrow=c(1,2), mar=c(4,2,4,2))
rhoB <- seq(1/min(eB), 1/max(eB), length.out=500)
plot(rhoB, sapply(rhoB, function(rho) sum(log(1 - rho*eB))), type="l", xlab=expression(rho), ylab="", main="Binary weights")
abline(v=1/range(eB), lty=2, lwd=2, col="#EB811B")
abline(h=0, v=0, lty=2, col="grey")
legend("bottomright", bty="n", legend=c(expression(plain(ln) * group("|", bold(I) - rho * bold(W), "|")), "extreme eigenvalues"), lty=c(1, 2), lwd=c(1, 2), col=c("black", "#EB811B"), cex=0.8)
rhoW <- seq(1/min(eW), 1/max(eW), length.out=500)
plot(rhoW, sapply(rhoW, function(rho) sum(log(1 - rho*eW))), type="l", xlab=expression(rho), ylab="", main="Row-standardised weights")
abline(v=1/range(eW), lty=2, lwd=2, col="#EB811B")
abline(h=0, v=0, lty=2, col="grey")
```

```{r}
par(oopar)
```

### APLE

```{r}
c(crossprod(eigW))
```
 
```{r}
W <- as(lwW, "CsparseMatrix")
sum(diag(W %*% W))
```
 
```{r}
var <- scale((wpom3$crimesrate), scale=FALSE)[,1]
set.seed(1)
aple.mc(var, lwW, nsim=999)
```
 
## Moran eigenvectors - spatial filtering
 
```{r}
x <- (wpom3$crimesrate)
moran.test(x, lwW)$estimate
```
 
```{r}
n <- nrow(wpom3)
M <- diag(n) - (tcrossprod(rep(1, n)))/n
MCM <- M %*% W %*% M
c((n/sum(W)) * crossprod(x, MCM %*% x)/crossprod(x, M %*% x))
```
 
```{r}
SF_pom <- SpatialFiltering(form_pom, data=wpom3, nb=pom_queen_nb, style="W", alpha=0.33, ExactEV = TRUE)
SF_pom
```
 
```{r}
SF_pom_lm <- lm(update(form_pom, . ~ . + fitted(SF_pom)), data=wpom3)
spdep::lm.morantest(SF_pom_lm, lwW)
```
 
### Extreme eigenvalues
 
```{r}
ecount(g1) <= 3*vcount(g1) - 6
```
 
```{r}
1/range(eB)
```
 
```{r}
rB <- lextrB(lwB)
1/c(rB)
```
 
 
```{r}
1/range(eW)
```
 
```{r}
rW <- lextrW(lwW)
1/c(rW)
```
 
### Alternatives
 
```{r}
B <- as(as(B0, "symmetricMatrix"), "unpackedMatrix")
arB <- eigs(B, 2, which = "BE")
1/rev(arB$values)
```
 
```{r}
W <- as(as(similar.listw(lwW), "CsparseMatrix"), "symmetricMatrix")
arW <- eigs(W, 2, which = "BE")
1/rev(arW$values)
```
```{r}
@article{A_Spatial_Analysis,
    author    = "Esther Akoto Amoako",
    title     = "A spatial analysis of robbery rate in the city of Detroit using exploratory data analysis approach",
    journal   = "International Cartographic Association",
    year      = "2021",
}
```












